Lossy compression of DNA sequencing quality data

### 2015-04-07 manuscript/q-compression.tex
Added placeholder plots.

@@

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the fragment library from
  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
  encoding. P32 and P256: Quality profile compression with 32 and 256
  profiles. R0-R5: Polynomial regression-based compression using 0-5
  degrees. Dashed line corresponds to lossless bzip compression of
  quality values. \em}\label{fig:mse_vs_bpbp_frag}
\end{figure}

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=3.65in]{mse_short.png}}
\caption{Mean squared error versus bits/base-call for for different
  compression methods applied to the short-jump library from
  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
  encoding. P32 and P256: Quality profile compression with 32 and 256
  profiles. R0-R5: Polynomial regression-based compression using 0-5
  degrees. Dashed line corresponds to lossless bzip compression of
  quality values. \em}\label{fig:mse_vs_bpbp_jump}
\end{figure}

@@

\begin{table*}[!tpb]
\centering
\caption[]{Rankings of different assemblies produced using proposed
  compression methods. Assemblies were constructed using
  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}
  2.4.1. data set provided by
  \textsc{gage}~\cite{Salzberg:2012rc}. Regression methods are sorted
  by their overall rank (determined by the sum of individual
  rankings).}
{\begin{tabular}{lcccccccc}
& overall & & & & & & \textsc{lap} & \textsc{lap} \\
compression            & rank & \textsc{snp}s & inversions & relocations & translocations & indels & fragment & short-jump \\
\hline
none                   & 1    & 5    & 1          & 1           & 1              & 1      & 1          & 1                    \\
profile (256)          & 3    & 2    & 1          & 3           & 3              & 4      & 2          & 4                    \\
profile (32)           & 4    & 3    & 1          & 4           & 6              & 5      & 3          & 3                    \\
pegression (4th order) & 5    & 6    & 1          & 6           & 5              & 2      & 7          & 8                    \\
regression (5th order) & 6    & 7    & 1          & 5           & 7              & 7      & 5          & 7                    \\
2-bin	               & 7    & 1    & 1          & 8           & 11             & 10     & 6          & 5                    \\
regression (3rd order) & 8    & 8    & 1          & 9           & 9              & 6      & 8          & 6                    \\
regression (2nd order) & 9    & 9    & 1          & 7           & 4              & 8      & 9          & 10                   \\
regression (1st order) & 10   & 10   & 11         & 10          & 8              & 9      & 10         & 9                    \\
regression (0 order) & 11   & 11   & 10         & 11          & 10             & 11     & 11         & 11
\\ \hline
\end{tabular}
}
\label{tab:assembly_ranks}
\end{table*}

@@

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files
  against \textit{Rhodobacter sphaeroides} reference genome using
  Bowtie2. Numbers corresponds to the proportion of mapped reads with
  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
  percentage of mapped reads by both the uncompressed and decompressed
  data. ``Uncompressed only'' denotes the percentage of reads mapped
  from the uncompressed data that are not mapped after
  decompression. ``Compressed only'' denotes the percentage of reads
  mapped from the decompressed data that were not mapped before
  compression.}
\begin{tabular}{llcccc}
    & & \multicolumn{4}{c}{percentage of reads mapped}\\
    \cline{3-6}
	library & compression strategy & total & shared & uncompressed only & compressed only \\ \hline
	& maximum quality & 77.88 & 77.88 & 22.12 & 0 \\
	&minimum quality & 100 & 100 & 0 & 0 \\
	&2-bin & 99.92 & 99.89 & 0.11 & 0.03 \\
	&regression (0 order) & 93.63 & 93.62 & 6.38 & 0.02 \\
	&regression (1st order) & 98.62 & 98.61 & 1.39 & 0.01 \\
	&regression (2nd order) & 99.42 & 99.39 & 0.61 & 0.03 \\
	fragment&regression (3rd order) & 99.56 & 99.53 & 0.47 & 0.03 \\
	&regression (4th order) & 99.71 & 99.68 & 0.32 & 0.03 \\ 
	&regression (5th order) & 99.79 & 99.75 & 0.25 & 0.03 \\
	&profile (32) & 99.9 & 99.86 & 0.14 & 0.04 \\ 
	&profile (256) & 99.95 & 99.92 & 0.08 & 0.03 \\

\\ \hline \\

	&maximum quality & 78.08 & 78.08 & 21.92 & 0 \\
	&minimum quality & 100 & 100 & 0 & 0 \\ 
	&2-bin & 97.38 & 97.38 & 2.62 & 0 \\
	&regression (0 order) & 96.07 & 94.11 & 5.89 & 1.96 \\
	&regression (1st order) & 99.24 & 98.98 & 1.02 & 0.26 \\ 
	&regression (2nd order) & 99.66 & 99.35 & 0.65 & 0.3 \\ 
	short-jump&regression (3rd order) & 99.94 & 99.61 & 0.39 & 0.33 \\ 
	&regression (4th order) & 99.99 & 99.64 & 0.36 & 0.35 \\ 
	&regression (5th order) & 100 & 99.67 & 0.33 & 0.33 \\ 
	&profile (32) & 99.74 & 99.44 & 0.56 & 0.3 \\
	&profile (256) & 99.76 & 99.52 & 0.48 & 0.24 \\
\end{tabular}

@@

\begin{table}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and
  decompressed \textsc{fastq} files. Numbers correspond to the
  proportion of bases/reads that are missed or gained after applying
  compression with respect to the uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lcccc}
	 & \multicolumn{2}{c}{fragments} & \multicolumn{2}{c}{short-Jump} \\
	 & \% bases & \% reads & \% bases & \% reads \\ \hline
	2-bin & 98.15 & 98.78 & 113.09 & 103.7 \\ 
	regression (0 order) & 86.71 & 62.64 & 43.31 & 27.51 \\
	regression (1st order) & 92.96 & 95.09 & 81.31 & 91.79 \\ 
	regression (2nd order) & 93.93 & 96.87 & 86.31 & 94.75 \\
	regression (3rd order) & 93.5 & 95.68 & 88.7 & 94.51 \\ 
	regression (4th order) & 94.43 & 96.7 & 91.94 & 97.02 \\ 
	regression (5th order) & 95.56 & 98.09 & 93.92 & 98.02 \\ 
	profile (32) & 99.12 & 98.96 & 102.89 & 102.24 \\
	profile (256) & 100 & 99.8 & 102.66 & 101.27 \\
\end{tabular}
\end{table}

>>>>

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{Mean squared error versus bits/base-call for different
  compression methods applied to the fragment library from
  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
  encoding. P32 and P256: Quality profile compression with 32 and 256
  profiles. R0-R5: Polynomial regression-based compression using 0-5
  degrees. Dashed line corresponds to lossless bzip compression of
  quality values. \em}\label{fig:mse_vs_bpbp_frag}
\end{figure*}

@@

%\begin{figure}[!tpb]%figure3
%\centerline{\includegraphics[width=3.65in]{mse_short.png}}
%\caption{Mean squared error versus bits/base-call for for different
%  compression methods applied to the short-jump library from
%  \textit{Rhodobacter sphaeroides} 2.4.1 data set. G/B: 2-bin
%  encoding. P32 and P256: Quality profile compression with 32 and 256
%  profiles. R0-R5: Polynomial regression-based compression using 0-5
%  degrees. Dashed line corresponds to lossless bzip compression of
%  quality values. \em}\label{fig:mse_vs_bpbp_jump}
%\end{figure}

@@

\begin{figure}[!tbp]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
\caption{Rankings of different assemblies produced using proposed
  compression methods. Assemblies were constructed using
  \textsc{allpaths-lg} and the \textit{Rhodobacter sphaeroides}
  2.4.1. data set provided by
  \textsc{gage}~\cite{Salzberg:2012rc}. Regression methods are sorted
  by their overall rank (determined by the sum of individual
  rankings). \em}
  \label{tab:assembly_ranks}
\end{figure}

@@

\begin{table*}[!tbhp]
\centering
\caption[]{Quality filtering/trimming for the uncompressed and
  decompressed \textsc{fastq} files. Numbers correspond to the
  number of reads that are kept or discarded after applying
  compression with respect to the original, uncompressed values.}
\label{tab:quality_control}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
  \cline{2-12}
& kept & 883576 &   0 &   0 & 883576 & 871035 & 12541 & 728349 & 155227 & 871510 & 12066 \\ 
{\em original} & discarded & 141858 &   0 &   0 & 141858 & 1882 & 139976 & 1475 & 140383 & 9214 & 132644 \\ 
\cline{2-12}
 & sum & 1025434 &   0 &   0 & 1025434 & 872917 & 152517 & 729824 & 295610 & 880724 & 144710 \\ 
\end{tabular}

\bigskip

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
 & & kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 872006 & 11570 & 878936 & 4640 & 877802 & 5774 & 876149 & 7427 & 877513 & 6063 \\ 
{\em original}  &  discarded & 10369 & 131489 & 10004 & 131854 & 8821 & 133037 & 15795 & 126063 & 17453 & 124405 \\
\cline{2-12}
&  sum & 882375 & 143059 & 888940 & 136494 & 886623 & 138811 & 891944 & 133490 & 894966 & 130468 \\
\end{tabular}

\bigskip

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
 & &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& kept & 881215 & 2361 & 878702 & 4874 & 879682 & 3894 & 881948 & 1628 & 881939 & 1637 \\ 
{\em original}  & discarded & 15985 & 125873 & 13434 & 128424 & 12855 & 129003 & 10872 & 130986 & 7179 & 134679 \\ 
\cline{2-12}
&  sum & 897200 & 128234 & 892136 & 133298 & 892537 & 132897 & 892820 & 132614 & 889118 & 136316 \\ 
\end{tabular}
\end{table*}

\begin{table*}[!tbhp]
\centering
\caption[]{Mapping results of decompressed \textsc{fastq} files
  against \textit{Rhodobacter sphaeroides} reference genome using
  Bowtie2. Numbers corresponds to the proportion of mapped reads with
  respect to the uncompressed \textsc{fastq}. ``Shared'' denotes the
  percentage of mapped reads by both the uncompressed and decompressed
  data. ``Uncompressed only'' denotes the percentage of reads mapped
  from the uncompressed data that are not mapped after
  decompression. ``Compressed only'' denotes the percentage of reads
  mapped from the decompressed data that were not mapped before
  compression.}
\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em MaxQual}} & \multicolumn{2}{c|}{{\em MinQual}} & \multicolumn{2}{c|}{{\em 2-bin}} & \multicolumn{2}{c|}{{\em Degree 0}} & \multicolumn{2}{c}{{\em Degree 1}} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 746716 & 145897 & 892613 &   0 & 891864 & 749 & 851682 & 40931 & 883390 & 9223 \\ 
{\em original} & unmapped &   0 & 132821 & 10821 & 122000 & 186 & 132635 &  67 & 132754 &  55 & 132766 \\ 
\cline{2-12}
& sum & 746716 & 278718 & 903434 & 122000 & 892050 & 133384 & 851749 & 173685 & 883445 & 141989 \\ 
\end{tabular}

\bigskip

\begin{tabular}{lr|cc|cc|cc|cc|cc}
 & & \multicolumn{2}{c|}{{\em Degree 3}} & \multicolumn{2}{c|}{{\em Degree 5}} & \multicolumn{2}{c|}{{\em Degree 7}} & \multicolumn{2}{c|}{{\em Profile (64)}} & \multicolumn{2}{c}{{\em Profile (128)}} \\
& & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped & mapped & unmapped \\ 
\cline{2-12}
& mapped & 889537 & 3076 & 891019 & 1594 & 891479 & 1134 & 891753 & 860 & 891952 & 661 \\ 
{\em original} & unmapped & 117 & 132704 & 155 & 132666 & 154 & 132667 & 144 & 132677 & 143 & 132678 \\ 
\cline{2-12}
& sum & 889654 & 135780 & 891174 & 134260 & 891633 & 133801 & 891897 & 133537 & 892095 & 133339 \\ 
\end{tabular}

\bigskip

\begin{tabular}{lr|cc|cc|cc|cc|cc}
&  & \multicolumn{2}{c|}{{\em Profile (256)}} & \multicolumn{2}{c|}{{\em QualComp (6)}} & \multicolumn{2}{c|}{{\em QualComp (10)}} & \multicolumn{2}{c|}{{\em QualComp (30)}} & \multicolumn{2}{c}{{\em QualComp (100)}} \\
& &  kept & discarded & kept & discarded & kept & discarded & kept & discarded & kept & discarded \\ 
\cline{2-12}
& mapped & 892051 & 562 & 891375 & 1238 & 891777 & 836 & 892233 & 380 & 892454 & 159 \\ 
{\em original}  & unmapped & 119 & 132702 & 304 & 132517 & 265 & 132556 & 220 & 132601 & 172 & 132649 \\ 
\cline{2-12}
& sum & 892170 & 133264 & 891679 & 133755 & 892042 & 133392 & 892453 & 132981 & 892626 & 132808 \\

\end{tabular}

@@
